\sheet

%>=====< Question 1 >=====<%

\question
What is the norm on $\Leb(X,Y)$? It satisfies two important equalities. Write and show them.

\subsection*{Solution}

\subsection{Norm on \texorpdfstring{$\Leb(X,Y)$}{the space of continuous operators}}
We can define the norm on $\Leb(X,Y)$ as follows:
\[ \| T \|_{\Leb} = \sup_{x\in X, \; \|x\|_X\leq1} \|T(x)\|_Y \]

%what are the conditions he is talking about

%>=====< Question 2 >=====<%

\question
Under which hypotheses on $X$ and $Y$ is $\Leb(X,Y)$ a Banach space?

\subsection*{Solution}

\subsection{When \texorpdfstring{$\Leb$}{the space of continuous operators} is a Banach space}
If $X$ is a normed space and $Y$ a Banach space then $\Leb(X,Y)$ is a Banach space.

%>=====< Question 3 >=====<%

\question
Let $\mathfrak{F} \subset \Leb(X,Y)$. For $\mathfrak{F}$ write the definition of pointwise and uniform boundedness. State and prove the UBP (or BS theorem).

\subsection*{Solution}

\subsection{Pointwise boundedness}
A family of operators $\mathfrak{F} \subset \Leb$ is pointwise bounded if and only if:
\[ \forall x \in X \; \exists M_x > 0: \quad \sup_{T\in\mathfrak{F}} \|T(x)\|_Y < M_x \]

\subsection{Uniform boundedness}
A family of operators $\mathfrak{F} \subset \Leb$ is uniformly bounded if and only if:
\[ \forall x \in X \; \exists M > 0: \quad \sup_{T\in\mathfrak{F}} \|T\|_{\Leb} < M \]
which is equal to writing:
\[ \forall x \in X \; \exists M > 0: \quad \sup_{T\in\mathfrak{F}} \sup_{\|x\|\leq1} < M \]

\subsection{Uniform boundedness Principle}
Let $X$,$Y$ be two Banach spaces, $\mathfrak{F}\subset\Leb(X,Y)$. If $\mathfrak{F}$ is pointwise bounded, then $\mathfrak{F}$ is also uniformly bounded.

\begin{proof}
    For all $n\in\N$ we define:
    \[ C_n \coloneqq \{ x\in X: \; \|T(x)\|_Y\leq n \; \forall T \in \mathfrak{F} \} \]
    Let us observe that $C_n$ is closed for all $n$. In fact, let $\seq{x}\subset C_n$ such that $x_n \xrightarrow{n\to\infty} x_0 \in X$, by continuity we have:
    \[ T(x_n) \xrightarrow{n\to\infty} T(x_0) \; \forall T \in \mathfrak{F} \implies \|T(x_n)\|_Y \xrightarrow{n\to\infty} \|T(x_0)\|_Y \]
    but since each $x_n \in C_n$ we also have:
    \begin{align*}
        & 0 \leq \|T(x_n)\|_Y \leq n \; \forall n \in \N \implies \|T(x_0)\|_Y \leq n \\
        & \implies x_0 \in C_n \implies X = \bigcup_{n=1}^\infty C_n
    \end{align*}
    Now we apply Baire's Theorem (\ref{Baire}) to write:
    \[ \exists n_0 \in \N \text{ such that } \int(C_{n_0}) \neq \emptyset \implies \exists \bar{B}_\epsilon(x_0) \subset C_{n_0} \]
    And so we have that:
    \[ \text{If } \|z\|_X \leq \epsilon \implies z+x_0 \in \bar{B}_\epsilon(x_0) \subset C_{n_0} \]
    \begin{align*}
        \implies \|T(z)\|_Y & \tikzmarknode{eq1}{=} \|T(z)+T(x_0)-T(x_0)\|_Y \\
        & \tikzmarknode{eq2}{\leq} \|T(z)+T(x_0)\|_Y + \|T(x_0)\|_Y \\
        & \tikzmarknode{eq3}{=} \|T(\underbrace{z+x_0}_{\in C_{n_0}})\|_Y + \|T(\underbrace{x_0}_{\in C_{n_0}})\|_Y \leq 2 n_0 \; \forall T \in \mathfrak{F}
    \end{align*}
\end{proof}\tikz[overlay,remember picture]{\draw[shorten >=1pt,shorten <=1pt] (eq1) -- (eq2) -- (eq3);}

%>=====< Question 4 >=====<%

\question
From the UBP it is possibile to infer an important property of operators defined by means of a point-wise limit. What is that? Justify your answer.

\subsection*{Solution}

%>=====< Question 5 >=====<%

\question
Write the definition of open mapping. Let $f: \R^n \to \R^m$ be continuous; under which extra hypotheses is $f$ open? Let $T: \R^n \to \R^n$ be linear and onto; why is $T$ open? State the OMT.

\subsection*{Solution}

%>=====< Question 6 >=====<%

\question
State and prove the IBM (or ICM) theorem. What is the analogous result in the finite dimensional case?

\subsection*{Solution}

%>=====< Question 7 >=====<%

\question
By the IBM theorem we can infer an important property about equivalent norms on Banach spaces. What is that? Justify your answer.

\subsection*{Solution}

%>=====< Question 8 >=====<%

\question
Write the definitions of: closed operator; graph of an operator. Show that an operator is linear and closed iff its graph is closed.

\subsection*{Solution}

%>=====< Question 9 >=====<%

\question
State and prove the closed graph theorem.

\subsection*{Solution}

%>=====< Question 10 >=====<%

\question
Write the definition of dual space.

\subsection*{Solution}

%>=====< Question 11 >=====<%

\question
Exhibit an example of $T \in (L^p)^{*}$. Compute $\|T\|_{*}$.

\subsection*{Solution}

%>=====< Question 12 >=====<%

\question
Let $(V, \langle \cdot, \cdot \rangle)$ be a finite dimensional vector space endowed with an inner product. Characterize $V^{*}$.

\subsection*{Solution}

%>=====< Question 13 >=====<%

\question
Let $Y$ be a vector subspace of $X = \R^2$. Given $f \in Y^{*}$ is it possibile to find $F \in X^{*}$ such that $F = f$ in $Y$ and $\|F\|_{X^{*}} = \|f\|_{Y^{*}}$ ?

\subsection*{Solution}

%>=====< Question 14 >=====<%

\question
State the Hahn-Banach theorem in the continuous extension form.

\subsection*{Solution}

%>=====< Question 15 >=====<%

\question
Let $X$ be a normed space, let $H$ be a hyperplane of $X$. When do we say that $H$ separates (or strictly separates) $A$ and $B$? If $X = \R^2$ , under which hypotheses on $A$ and $B$ is it possibile to find a line $H$ which separates them?

\subsection*{Solution}

%>=====< Question 16 >=====<%

\question
State the Hahn-Banach theorem in the separation form (first and second version).

\subsection*{Solution}
